---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2024"
subtitle: "Assignment 3 - Due date 02/01/24"
author: "Jenn McNeill"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---
```{r}

```
  
## Directions
  
You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github. 

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A02_Sp24.Rmd"). Then change "Student Name" on line 4 with your name. 

Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

Please keep this R code chunk options for the report. It is easier for us to grade when we can see code and output together. And the tidy.opts will make sure that line breaks on your code chunks are automatically added for better visualization. 

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=FALSE) 
```

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai. 

## Questions 

Consider the same data you used for A2 from the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx". The data comes from the US Energy Information and Administration and corresponds to the December 2022 **Monthly** Energy Review. Once again you will work only with the following columns: Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption. Create a data frame structure with these three time series only. 

R packages needed for this assignment:"forecast","tseries", and "Kendall". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.\\

```{r}
#Load/install required package here

library(forecast)
library(tseries)
library(Kendall)
library(lubridate)
library(ggplot2)
library(cowplot)
library(dplyr)
getwd()
```


##Trend Component

### Q1
For each time series, i.e., Renewable Energy Production and Hydroelectric Consumption create three plots: one with time series, one with the ACF and with the PACF. You may use the some code form A2, but I want all the three plots side by side as in a grid.
(Hint: use function `plot_grid()` from the `cowplot` package)

```{r Q1}
#Importing data set
renewable_energy_raw <- read.table(file="./Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.csv", header=TRUE,dec = ".",sep=",",stringsAsFactors = TRUE) 

#Select columns of interest
renewables <- select(renewable_energy_raw, Month, Total.Biomass.Energy.Production:Hydroelectric.Power.Consumption)

#Create time series objects
ts_rp <- ts(renewables[,3],start=c(1973,1), frequency=12)
ts_hc <- ts(renewables[,4],start=c(1973,1), frequency=12)

#Compute Acf for each time series
rp_acf = ggAcf(ts_rp,lag.max=40,plot=TRUE) + ggtitle(NULL)
hc_acf = ggAcf(ts_hc,lag.max=40,plot=TRUE) + ggtitle(NULL)

#Compute Pacf for each time series
rp_pacf = ggPacf(ts_rp,lag.max=40,plot=TRUE) + ggtitle(NULL)
hc_pacf = ggPacf(ts_hc,lag.max=40,plot=TRUE) + ggtitle(NULL)

#Plot the time series objects
rp_ts_plot <-
  autoplot(ts_rp)+
  xlab("Year")+
  ylab("Yearly Renewable Production in Trillion Btu")
rp_ts_plot

hc_ts_plot <-
  autoplot(ts_hc)+
  xlab("Year")+
  ylab("Yearly Hydroelectric Consumption in Trillion Btu")
hc_ts_plot 

#Plot all three plots side by side
plot_grid(rp_ts_plot,
          rp_acf,
          rp_pacf,
          ncol = 3, align = "v")

plot_grid(hc_ts_plot,
          hc_acf,
          hc_pacf,
          ncol = 3, align = "v")
```

### Q2
From the plot in Q1, do the series Total Renewable Energy Production and Hydroelectric Power Consumption appear to have a trend? If yes, what kind of trend?

### Q3
Use the *lm()* function to fit a linear trend to the three time series. Ask R to print the summary of the regression. Interpret the regression output, i.e., slope and intercept. Save the regression coefficients for further analysis.

```{r Q3}
#Create variables to use for indexing
t <- 1:nobs

#Fit a linear trend to the renewable production time series
rp_lm <- lm(renewables[,3]~t)
summary(rp_lm)
#Store slope and intercept coefficients
rp_beta0 <- rp_lm$coefficients[1]
rp_beta1 <- rp_lm$coefficients[2]

#Fit a linear trend to the hydroelectric consumption time series 
hc_lm <- lm(renewables[,4]~t)
summary(hc_lm)
#Store slope and intercept coefficients
hc_beta0 <- hc_lm$coefficients[1]
hc_beta1 <- hc_lm$coefficients[2]

rp_beta0
rp_beta1
hc_beta0
hc_beta1
```
 
### Q4
Use the regression coefficients from Q3 to detrend the series. Plot the detrended series and compare with the plots from Q1. What happened? Did anything change?

```{r Q4}
ts_rp_detrend <- ts((renewables[,3] - (rp_beta0 + (rp_beta1*t))),
                         start=c(1973,1), frequency=12)

autoplot(ts_rp, color = "slateblue4")+
  autolayer(ts_rp_detrend, color = "tomato3")+
  xlab("Year")+
  ylab("Yearly Renewable Energy Production in Trillion Btu")+
  theme_light()+
  theme(legend.position = "none")

ts_hc_detrend <- ts((renewables[,4] - (hc_beta0 + (hc_beta1*t))),
                         start=c(1973,1), frequency=12)

autoplot(ts_hc, color = "palegreen4")+
  autolayer(ts_hc_detrend, color = "orchid3")+
  xlab("Year")+
  ylab("Yearly Hydroelectric Power Consumption in Trillion Btu")+
  theme_light()+
  theme(legend.position = "none")
```


### Q5
Plot ACF and PACF for the detrended series and compare with the plots from Q1. You may use plot_grid() again to get them side by side. nut mot mandatory. Did the plots change? How?

```{r Q5}
#Compare the Acf for renewable production with and without trend
par(mfrow=c(1,2))
  Acf(ts_rp,lag.max=40,plot=TRUE) 
  Acf(ts_rp_detrend,lag.max=40,plot=TRUE)

#Compare the Pacf for renewable production with and without trend 
par(mfrow=c(1,2))
  Pacf(ts_rp,lag.max=40,plot=TRUE) 
  Pacf(ts_rp_detrend,lag.max=40,plot=TRUE)

#Compare the Acf for hydroelectric consumption with and without trend
par(mfrow=c(1,2))
  Acf(ts_hc,lag.max=40,plot=TRUE) 
  Acf(ts_hc_detrend,lag.max=40,plot=TRUE)

#Compare the Pacf for hydroelectric consumption with and without trend  
par(mfrow=c(1,2))
  Pacf(ts_hc,lag.max=40,plot=TRUE) 
  Pacf(ts_hc_detrend,lag.max=40,plot=TRUE)
```


## Seasonal Component

Set aside the detrended series and consider the original series again from Q1 to answer Q6 to Q8.

### Q6
Just by looking at the time series and the acf plots, do the series seem to have a seasonal trend? No need to run any code to answer your question. Just type in you answer below.

### Q7
Use function *lm()* to fit a seasonal means model (i.e. using the seasonal dummies) the two time series. Ask R to print the summary of the regression. Interpret the regression output. From the results which series have a seasonal trend? Do the results match you answer to Q6?

```{r Q7}


```


### Q8
Use the regression coefficients from Q7 to deseason the series. Plot the deseason series and compare with the plots from part Q1. Did anything change?

### Q9
Plot ACF and PACF for the deseason series and compare with the plots from Q1. You may use plot_grid() again to get them side by side. nut mot mandatory. Did the plots change? How?


